---
title: "Statewide New Diagnoses"
author: "Erich Denk"
date: "4/11/2019"
output: pdf_document
---

AidsVu provides state diagnoses data for the years 2008 to 2016. Let's upload that data so we can combine it to create a longitudinal dataset. We will need to add the relevant year before comibing as well.

```{r}
require(tidyverse)
require(readxl)

state_diag_2008 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2008.xlsx")
state_diag_2009 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2009.xlsx")
state_diag_2010 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2010.xlsx")
state_diag_2011 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2011.xlsx")
state_diag_2012 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2012.xlsx")
state_diag_2013 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2013.xlsx")
state_diag_2014 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2014.xlsx")
state_diag_2015 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2015.xlsx")
state_diag_2016 <- read_excel("Data/AIDSVu_State_New-DX-Data-Sets_2016.xlsx")
```

```{r}
varnames <- state_diag_2008[2,]

state_diag_2008 <- state_diag_2008%>%
  mutate(year = 2008)%>%
  slice(2:53)
state_diag_2009 <- state_diag_2009%>%
  mutate(year = 2009)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" =`2009 State New Diagnoses Data - AIDSVu`)
state_diag_2010 <- state_diag_2010%>%
  mutate(year = 2010)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2010 State New Diagnoses Data - AIDSVu`)
state_diag_2011 <- state_diag_2011%>%
  mutate(year = 2011)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2011 State New Diagnoses Data - AIDSVu`)
state_diag_2012 <- state_diag_2012%>%
  mutate(year = 2012)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2012 State New Diagnoses Data - AIDSVu`)
state_diag_2013 <- state_diag_2013%>%
  mutate(year = 2013)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2013 State New Diagnoses Data - AIDSVu`)
state_diag_2014 <- state_diag_2014%>%
  mutate(year = 2014)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2014 State New Diagnoses Data - AIDSVu`)
state_diag_2015 <- state_diag_2015%>%
  mutate(year = 2015)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2015 State New Diagnoses Data - AIDSVu`)
state_diag_2016 <- state_diag_2016%>%
  mutate(year = 2016)%>%
  slice(2:53)%>%
  rename("2008 State New Diagnoses Data -AIDSVu" = `2016 State New Diagnoses Data - AIDSVu`)
```

```{r}

diag_08_16 <- 
  bind_rows(state_diag_2008,state_diag_2009)%>%
  bind_rows(.,state_diag_2010)%>%
  bind_rows(.,state_diag_2011)%>%
  bind_rows(.,state_diag_2012)%>%
  bind_rows(.,state_diag_2013)%>%
  bind_rows(.,state_diag_2014)%>%
  bind_rows(.,state_diag_2015)%>%
  bind_rows(.,state_diag_2016)
head(diag_08_16)
  
```

We need to sort through some of what we need. First, let's get real variable names and also just select some of what we need
```{r}
colnames(diag_08_16) <-  c(varnames, "Year1", "drop")
```
```{r}
diag_08_16 <- diag_08_16%>%
  slice(2:468)%>%
  select(Year1, `State Abbreviation`, State, 
           `New Diagnoses State Rate`,  `New Diagnoses State Cases`,
           `New Diagnoses IDU Total Cases`)
head(diag_08_16)
```
I think a variable that shows the percentage of new diagnoses that are from IDUs would be helpful.
```{r}
diag_08_16 <- diag_08_16%>%
  mutate(per_IDU = (as.numeric(`New Diagnoses IDU Total Cases`)/
                      as.numeric(`New Diagnoses State Cases`))*100)
```

Let's save what we have as a csv in our data folder.
```{r}
write_csv(diag_08_16, 
          file.path("Data/statewide-new-diagnoses-clean.csv"))
```


