---
title: "PPOL 670 Project Showcase"
author: "Ivy Deng, Erich Denk, and Ann-Marie Stringer"
date: "4/29/2019"
output: slidy_presentation
footer: "            Team AMIE            "
incremental: TRUE
duration: 12
theme: united
highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(tidyverse)
require(caret)
require(recipes)
require(pdp)
require(here)
```

## <span style="color:steelblue">**Problem Statement and Background** </span>

- HIV still a problem in the US today more than a million americans live with hiv (CDC)
- No known cure to HIV, only ARV
- HIV disproportionately affects historically marginalized communities
    - MSM population
    - Injection Drug Users 
    - Black population
- Two prevention policies
  1. Syringe Services Programs (SSPs)/ Syringe Exchange Programs (SEPs)
  2. Pre-exposure prophylaxis (PrEP)
  
## <span style="color:steelblue">**Problem Statement and Background** </span>
```{r, include=FALSE}
idu_dat <- read_csv("idu_dat.csv")

idu_plot <- 
  idu_dat%>%
  ggplot(mapping = aes(x = as.character(idu_dat$Year),
                       y = idu_dat$New.Diagnoses.IDU.Total.Cases))+
  geom_bar(stat ="identity", fill = "#008FD5", alpha = 0.8)+
    ggthemes::theme_fivethirtyeight()+
    theme(panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(face = "italic", size =10),
        plot.caption = element_text(face = "italic"))+
    labs(title = "The Needle and the Damage Done",
         subtitle = "HIV cases transmitted by Injection Drug Use are falling",
         caption = "Source: AidsVu")
```

*Two trends that brought our research question about*
```{r, include=FALSE}
SEP_plot <- idu_dat%>%
  mutate(Yes = if_else(Allowed == "Yes", 1, 0))%>%
  ggplot(aes(x = as.character(Year), y = Yes))+
  geom_bar(stat = "identity",fill = "#FF2700", alpha = 0.8)+
  ggthemes::theme_fivethirtyeight()+
  theme(panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(face = "italic", size =10),
        plot.caption = element_text(face = "italic"),)+
  labs(title = "States increasingly allowing \nSyringe Exchange Programs",
       subtitle = "SEPs, while controversial, are becoming a common policy tool",
       caption = "Source: Center for Disease Control")
```
```{r, fig.width=9, fig.height=4, fig.align='center'}
plots <- grid.arrange(idu_plot,SEP_plot, ncol = 2)
```

## <span style="color:steelblue">**Aim** </span>

 >Build a dashboard that walks policy makers and key stakeholders through the unmet need of these two prevention methods (PrEP and SSP). And, for the states that do allow SSPs, can we predict when they will allow them using new diagnoses data?

## <span style="color:steelblue">**Approach and Design** </span>
- Three main data sources
  1. AidsVu
  2. Center for Disease Control and Prevention State Laws
  3. NASEN SSP locations
  
## <span style="color:steelblue">**Approach and Design (cont)** </span>
- Our original goal -- countywide incidence - yearly incidence data not available

- Time series at state level doesn’t really show visual change in mapping  - not really impactful. 

- The methods we did use..
  - Web scraping (very minimal, just NASEN)
  - Data Wrangling
  - Data Visualization
  - Mapping
  - Machine Learning
  - Shiny

## <span style="color:steelblue">**Methods and Tools** </span>

- The Shiny App
- Urban Maps
- Caret
- ggiraph

## <span style="color:steelblue">**Results** </span> 

Our dashboard: *link to dashboard*

## <span style="color:steelblue">**Results** </span>

The machine learning outcomes
insert ggiraph
```{r}

```

## <span style="color:steelblue">**Results** </span>

## <span style="color:steelblue">**Lessons Learned** </span>

- Be flexible! Try to let your preconcieved notions go

## <span style="color:steelblue">**Thanks!** </span>



