---
title: "PPOL 670 Project Showcase"
author: "Ivy Deng, Erich Denk, and Ann-Marie Stringer"
date: "4/29/2019"
output:
  ioslides_presentation: default
  slidy_presentation: default
incremental: yes
footer: '            Team AMIE            '
duration: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(tidyverse)
require(caret)
require(recipes)
require(pdp)
require(here)
require(ggplot2)
require(ggthemes)
require(ggiraph)
```

# <span style="color:steelblue">**Problem Statement and Background** </span>

- HIV still a problem in the US today more than a million americans live with hiv (CDC)
- No known cure to HIV, only ARV
- HIV disproportionately affects historically marginalized communities
    - MSM population
    - Injection Drug Users 
    - Black population
    - Sex Workers
- Two prevention policies
  1. Syringe Services Programs (SSPs)/ Syringe Exchange Programs (SEPs)
  2. Pre-exposure prophylaxis (PrEP)
  
# <span style="color:steelblue">**Problem Statement and Background** </span>
```{r, include=FALSE}
idu_dat <- read_csv("idu_dat.csv")

idu_plot <- 
  idu_dat%>%
  ggplot(mapping = aes(x = as.character(idu_dat$Year),
                       y = idu_dat$New.Diagnoses.IDU.Total.Cases))+
  geom_bar(stat ="identity", fill = "#008FD5", alpha = 0.8)+
    ggthemes::theme_fivethirtyeight()+
    theme(panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(face = "italic", size =10),
        plot.caption = element_text(face = "italic"))+
    labs(title = "The Needle and the Damage Done",
         subtitle = "HIV cases transmitted by Injection Drug Use are falling",
         caption = "Source: AidsVu")
```

*Two trends that brought our research question about*
```{r, include=FALSE}
SEP_plot <- idu_dat%>%
  mutate(Yes = if_else(Allowed == "Yes", 1, 0))%>%
  ggplot(aes(x = as.character(Year), y = Yes))+
  geom_bar(stat = "identity",fill = "#FF2700", alpha = 0.8)+
  ggthemes::theme_fivethirtyeight()+
  theme(panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(face = "italic", size =10),
        plot.caption = element_text(face = "italic")) +
  labs(title = "States increasingly allowing \nSyringe Exchange Programs",
       subtitle = "SEPs, while controversial, are becoming a common policy tool",
       caption = "Source: Center for Disease Control")
```
```{r, fig.width=9, fig.height=4, fig.align='center'}
plots <- grid.arrange(idu_plot,SEP_plot, ncol = 2)
```

# <span style="color:steelblue">**Aim** </span>

 >Build a dashboard that walks policy makers and key stakeholders through the unmet need of these two prevention methods (PrEP and SSP). And, for the states that do allow SSPs, can we predict when they will allow them using new diagnoses data?

# <span style="color:steelblue">**Approach and Design** </span>
- Three main data sources
  1. AidsVu
  2. Center for Disease Control and Prevention State Laws
  3. NASEN SSP locations
  
# <span style="color:steelblue">**Approach and Design (cont)** </span>
- Our original goal -- countywide incidence - yearly incidence data not available

- Time series at state level doesn’t really show visual change in mapping  - not really impactful. 

- The methods we did use..
  - Web scraping (just NASEN)
  - Data Wrangling
  - Data Visualization
  - Mapping
  - Machine Learning
  - Shiny

# <span style="color:steelblue">**Methods and Tools** </span>

- Dplyr and the <span style="color:dodgerblue">**tidyverse** </span>
- The Shiny App
- Urban Maps
- Caret
- ggiraph
- ggplot2

# <span style="color:steelblue">**Results** </span> 

The shiny app dashboard. Combines a number of efforts together

<<<<<<< HEAD
<span style="color:dodgerblue"> Our dashboard: </span> 
https://ivy-deng.shinyapps.io/Shiny-Apps/

- Unmet PrEP Demand in states such as Texas, California, Louisiana, Georgia, and Florida. California, Texas, and Florida show a decreasing trend of number of PrEP users overtime. 

- Southern states in general have more cases of HIV new diagnosis than other regions. However, there are fewer users of PrEP in those states compared to other regions. 

-(Ann Marie: Results from HIV IDU and SSP map)
=======
 Our [dashboard](https://ivy-deng.shinyapps.io/Shiny-Apps/)
>>>>>>> f09dd1c093d5c13292ea068eb750521db0a5dee6

# <span style="color:steelblue">**Results** </span>
The second (an less successful) effort: Our supervised machine learning results

- Random regression trees
- knn
- random forests

Extremely good at predicting which state-years would *not* allow Syringe Exchange Programs, but not very good at predicting those that do.

Random Forest with extratrees and mtry = 34 was the best model. In that model, here is the relative importance

# <span style="color:steelblue">**Results** </span>
```{r, include = FALSE}
require(wesanderson)
require(ggplot2)
require(ggiraph)
require(ggthemes)

pal1 <- wes_palette(name = "GrandBudapest2", 6, type = "continuous")
grab_varimp <- read_csv(here("R-markdown-Files/ggiraph.csv"))
```
```{r, fig.width=5,fig.height=5, fig.align='center'}
plot_varimp <- 
  grab_varimp %>% 
    mutate(keyword = fct_reorder(keyword,importance)) %>% 
    mutate(sum_report = str_glue('keyword: "{keyword}"\nImportance: {round(importance,2)}%')) %>% 
    top_n(10, importance) %>% 
    ggplot(aes(keyword,importance,
               fill=importance)) +
    geom_point() +
    geom_hline(aes(yintercept = max(importance)),color="grey20",size=1,alpha=.25) +
    geom_bar_interactive (aes(tooltip=sum_report),
                         stat="identity",color="grey10") + 
    coord_flip() +
    scale_fill_gradientn(colors = pal1) +
    theme_hc() +
    labs(y="Importance of the Feature in Predicting SEPs\n(higher values == more important)",
         x="Features")+
    theme(legend.position = "none",
          axis.title = element_text(family="serif",size=10),
          axis.text = element_text(family="serif",size=10)) -> plt_obj
  interactive_plot <- girafe(ggobj = plt_obj,width = 4,height=4)
  girafe_options(interactive_plot, opts_toolbar(position = "topleft"))

```

<<<<<<< HEAD

=======
>>>>>>> f09dd1c093d5c13292ea068eb750521db0a5dee6
# <span style="color:steelblue">**Lessons Learned** </span>

- Be flexible! Try to let your preconcieved notions go

- Opportunity to learn about a new topic

- [urbnmapr] package! Perfect for mapping geospatial data. 

- Shiny app. Efficient way to display data in various forms and let audience interact with data. 


# <span style="color:steelblue">**Conclusions** </span>

- States with a high HIV prevalence should invest more resources in supporting and promoting HIV prevention programs (PrEP and SSPs)

- 

# <span style="color:steelblue">**Thanks!** </span>

- Any questions?

