---
title: "PPOL 670 Project Showcase"
author: "Team AMIE"
date: "4/25/2019"
output: slidy_presentation
footer: "Team AMIE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(tidyverse)
require(caret)
require(recipes)
require(pdp)
require(here)
```

## Problem Statement and Background

- HIV is transmitted through a variety of ways. One is Injection Drug Use (IDU)
- Some states have attempted to address the tranmission of HIV and other disease via Syringe Exchange Programs (SEPs)
- Another policy intervention is encouragement of PrEP, a treatment for at-risk people before they come into contact with HIV
- We sought to begin to understand how well these policy interventions work

## Problem Statement and Background (cont)

```{r}
idu_dat <- read_csv("idu_dat.csv")

idu_plot <- 
  idu_dat%>%
  ggplot(mapping = aes(x = idu_dat$Year,
                       y = idu_dat$New.Diagnoses.IDU.Total.Cases))+
  geom_bar(stat ="identity", fill = "#008FD5", alpha = 0.8)+
    ggthemes::theme_fivethirtyeight()+
    theme(panel.background = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(face = "italic", size =10),
        plot.caption = element_text(face = "italic"))+
    labs(title = "The Needle and the Damage Done",
         subtitle = "HIV cases transmitted by Injection Drug Use are falling",
         caption = "Source: AidsVu")
```

```{r, idu_plot, fig.align='center', fig.height=5, fig.width=6}
idu_plot
```
## Approach/Design

## Approach/Design (cont)

## Methods/Tools

## Methods/Tools

## Results

## Lessons Learned

## Thanks

